---
title: "Machine Learning Workflow: Adding Nonlinearities to Predictive Modeling"
subtitle: "Biostat 212A"
author: "Dr. Jin Zhou @ UCLA"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

Display system information for reproducibility.

::: {.panel-tabset}

#### Python

```{python}
import IPython
print(IPython.sys_info())
```

#### R

```{r}
sessionInfo()
```

:::

## Overview

![](https://www.tidymodels.org/start/resampling/img/resampling.svg)


We illustrate the typical machine learning workflow for regression problems using the `Wage` data set from R `ISLR2` package. The steps are

1. Initial splitting to test and non-test sets.

2. Pre-processing of data: one-hot-encoder for categorical variables, add nonlinear features for some continuous predictors.

3. Choose a learner/method. Elastic net in this example.

4. Tune the hyper-parameter(s) (`alpha` and `l1_ratio` in elastic net) using $K$-fold cross-validation (CV) on the non-test data.

5. Choose the best model by CV and refit it on the whole non-test data.

6. Final prediction on the test data.

These steps completes the process of training and evaluating **one** machine learning method. We repeat the same process for other learners, e.g., random forest or neural network, using the same test/non-test and CV split. The final report compares the learners based on CV and test errors.

## Wage data


A documentation of the `Wage` data is [here](https://www.rdocumentation.org/packages/ISLR2/versions/1.3-2/topics/Wage). The goal is to predict the `wage`.

::: {.panel-tabset}

#### R

```{r}
library(GGally)
library(ISLR2)
library(tidymodels)
library(tidyverse)

Wage <- as_tibble(Wage) %>%
  select(-region) %>%
  print(width = Inf)

# Numerical summaries
summary(Wage)
```

Graphical summary takes longer to run so suppressed here.
```{r}
#| eval: false
# Graphical summaries
ggpairs(
  data = Wage, 
  mapping = aes(alpha = 0.25), 
  lower = list(continuous = "smooth")
  ) + 
  labs(title = "Wage Data")
```

#### Python

```{python}
# Load the pandas library
import pandas as pd
# Load numpy for array manipulation
import numpy as np
# Load seaborn plotting library
import seaborn as sns
import matplotlib.pyplot as plt

# Set font sizes in plots
sns.set(font_scale = 1.2)
# Display all columns
pd.set_option('display.max_columns', None)

Wage = pd.read_csv("../data/Wage.csv").drop(['logwage', 'region'], axis = 1)
Wage
```

```{python}
# Numerical summaries
Wage.describe()
```
Graphical summary takes longer to run so suppressed here.
```{python}
#| eval: true
# Graphical summaries
plt.figure()
sns.pairplot(data = Wage);
plt.show()
```



:::


## Initial split into test and non-test sets

::: {.panel-tabset}



#### R

```{r}
# For reproducibility
set.seed(425)
data_split <- initial_split(
  Wage, 
  # # stratify by percentiles
  # strata = "Salary", 
  prop = 0.75
  )

Wage_other <- training(data_split)
dim(Wage_other)
Wage_test <- testing(data_split)
dim(Wage_test)
```

#### Python

```{python}
from sklearn.model_selection import train_test_split

Wage_other, Wage_test = train_test_split(
  Wage, 
  train_size = 0.75,
  random_state = 425, # seed
  )
Wage_test.shape
Wage_other.shape
```

Separate $X$ and $y$.
```{python}
# Non-test X and y
X_other = Wage_other.drop(['wage'], axis = 1)
y_other = Wage_other.wage
# Test X and y
X_test = Wage_test.drop(['wage'], axis = 1)
y_test = Wage_test.wage
```

:::

## Preprocessing (Python) or recipe (R)

For regularization methods such as ridge and lasso, it is essential to center and scale predictors. 

::: {.panel-tabset}



#### R

```{r}
#| eval: true
norm_recipe <- 
  recipe(
    wage ~ ., 
    data = Wage_other
  ) %>%
  # create traditional dummy variables
  step_dummy(all_nominal()) %>%
  # zero-variance filter
  step_zv(all_predictors()) %>% 
  # B-splines of age
  step_bs(age, deg_free = 5) %>%
  # B-splines of year
  step_bs(year, deg_free = 4) %>%
  # center and scale numeric data
  step_normalize(all_predictors()) %>%
  # estimate the means and standard deviations
  prep(training = Wage_other, retain = TRUE)
norm_recipe
```

#### Python

Pre-processor for one-hot coding of categorical variables and then standardizing all numeric predictors.
```{python}
from sklearn.preprocessing import OneHotEncoder, StandardScaler, SplineTransformer
from sklearn.compose import make_column_transformer
from sklearn.pipeline import Pipeline


col_tf = make_column_transformer(
  # OHE transformer for categorical variables
  (OneHotEncoder(drop = 'first'), ['maritl', 'race', 'education', 'jobclass', 'health', 'health_ins']),
  # Nonlinear features by splines of age and year
  (SplineTransformer(
    degree = 3,
    n_knots = 5,
    extrapolation = 'linear'
    ), ['age']),
  (SplineTransformer(
    degree = 3,
    n_knots = 4,
    extrapolation = 'linear'
    ), ['year']),
  remainder = 'passthrough'
)
# Standardization transformer
std_tf = StandardScaler()
```

:::

## Model

::: {.panel-tabset}

We use elastic net in this example.



#### R

```{r}
#| eval: true
enet_mod <- 
  # mixture = 0 (ridge), mixture = 1 (lasso)
  linear_reg(penalty = tune(), mixture = tune()) %>% 
  set_engine("glmnet")
enet_mod
```

#### Python

```{python}
from sklearn.linear_model import ElasticNet

enet_mod = ElasticNet()
enet_mod
```

:::

## Pipeline (Python) or workflow (R)

Here we bundle the preprocessing step (Python) or recipe (R) and model.

::: {.panel-tabset}



#### R

```{r}
#| eval: true
lr_wf <- 
  workflow() %>%
  add_model(enet_mod) %>%
  add_recipe(norm_recipe)
lr_wf
```

#### Python

```{python}
pipe = Pipeline(steps = [
  ("col_tf", col_tf),
  ("std_tf", std_tf),
  ("model", enet_mod)
  ])
pipe
```

:::

## Tuning grid

Set up the 2D grid for tuning.

::: {.panel-tabset}



#### R

```{r}
#| eval: true
param_grid <-grid_regular(
  penalty(range = c(-5, 0), trans = log10_trans()), 
  mixture(range = c(0, 1)),
  levels = c(penalty = 50, mixture = 6)
  )
param_grid
```

#### Python

```{python}
# Tune hyper-parameter(s)
alphas = np.logspace(start = -6, stop = -1, base = 10, num = 50)
l1_ratios = np.linspace(start = 0, stop = 1, num = 6)
# n_knots = [2, 3, 4, 5]
tuned_parameters = {
  "model__alpha": alphas,
  "model__l1_ratio": l1_ratios
  # "bs_tf__n_knots": n_knots
  }
tuned_parameters  
```

:::

## Cross-validation (CV)

::: {.panel-tabset}



#### R

Set cross-validation partitions.
```{r}
#| eval: true
set.seed(250)
folds <- vfold_cv(Wage_other, v = 10)
folds
```

Fit cross-validation.
```{r}
#| eval: true
enet_fit <- 
  lr_wf %>%
  tune_grid(
    resamples = folds,
    grid = param_grid,
    )
enet_fit
```

Visualize CV criterion.
```{r}
#| eval: true
enet_fit %>%
  collect_metrics() %>%
  print(width = Inf) %>%
  filter(.metric == "rmse") %>%
  ggplot(mapping = aes(x = penalty, y = mean)) + 
  geom_point() + 
  geom_line(aes(group = mixture)) + 
  labs(x = "Penalty", y = "CV RMSE") + 
  scale_x_log10(labels = scales::label_number())
```

Show the top 5 models ($\lambda$ values)
```{r}
#| eval: true
enet_fit %>%
  show_best("rmse")
```

Let's select the best model
```{r}
#| eval: true
best_enet <- enet_fit %>%
  select_best("rmse")
best_enet
```

#### Python

Set up CV partitions and CV criterion.
```{python}
from sklearn.model_selection import GridSearchCV

# Set up CV
n_folds = 10
search = GridSearchCV(
  pipe,
  tuned_parameters,
  cv = n_folds, 
  scoring = "neg_root_mean_squared_error",
  # Refit the best model on the whole data set
  refit = True
  )
```

Fit CV. This is typically the most time-consuming step.
```{python}
# Fit CV
search.fit(X_other, y_other)
```

Visualize CV results (TODO).
```{python}
#| eval: false
#| code-fold: true
cv_res = pd.DataFrame({
  "alpha": alphas,
  "rmse": -search.cv_results_["mean_test_score"]
  })

plt.figure()
sns.relplot(
  data = cv_res,
  x = "alpha",
  y = "rmse"
  ).set(
    xlabel = "alpha",
    ylabel = "CV RMSE",
    xscale = "log"
);
plt.show()
```

Best CV RMSE:
```{python}
-search.best_score_
```

:::

## Finalize our model

Now we are done tuning. Finally, let’s fit this final model to the whole training data and use our test data to estimate the model performance we expect to see with new data.

::: {.panel-tabset}




#### R

```{r}
#| eval: true
# Final workflow
final_wf <- lr_wf %>%
  finalize_workflow(best_enet)
final_wf

# Fit the whole training set, then predict the test cases
final_fit <- 
  final_wf %>%
  last_fit(data_split)
final_fit

# Test metrics
final_fit %>% collect_metrics()
```

#### Python

Since we called `GridSearchCV` with `refit = True`, the best model fit on the whole non-test data is readily available.
```{python}
search.best_estimator_
```

The final prediction RMSE on the test set is
```{python}
from sklearn.metrics import mean_squared_error

mean_squared_error(y_test, search.best_estimator_.predict(X_test), squared = False)
```

:::