---
title: "Machine Learning Workflow: Classifiers With Nonlinear Features"
subtitle: "Biostat 212A"
author: "Dr. Jin Zhou @ UCLA"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

Display system information for reproducibility.

::: {.panel-tabset}

#### R

```{r}
sessionInfo()
```

#### Python

```{python}
import IPython
print(IPython.sys_info())
```

:::

## Overview

![](https://www.tidymodels.org/start/resampling/img/resampling.svg)


We illustrate the typical machine learning workflow for regression problems using the `Default` data set from R `ISLR2` package. Our goal is to classify whether a credit card customer will default or not. The steps are

1. Initial splitting to test and non-test sets.

2. Pre-processing of data: one-hot-encoder for categorical variables, add **nonlinear features** (B-splines) for some continuous predictors.

3. Choose **a set of candidate classifiers**: logistic regression, LDA, QDA, NB, KNN.

4. Tune the hyper-parameter(s) (`n_knots` in SplineTransformer, classifier) using 10-fold cross-validation (CV) on the non-test data.

5. Choose the best model by CV and refit it on the whole non-test data.

6. Final classification on the test data.

## Default data

A documentation of the `Default` data is [here](https://www.rdocumentation.org/packages/ISLR2/versions/1.3-2/topics/Default). The goal is to classify whether a credit card customer will default or not.

::: {.panel-tabset}

#### R

```{r}
##| eval: true
library(GGally)
library(ISLR2)
library(tidymodels)
library(tidyverse)

Default <- as_tibble(Default) %>%
  print(width = Inf)

# Numerical summaries
summary(Default)
```

Graphical summary.
```{r}
#| eval: false
# Graphical summaries
ggpairs(
  data = Default, 
  mapping = aes(alpha = 0.25), 
  lower = list(continuous = "smooth")
  ) + 
  labs(title = "Default Data")
```

#### Python

```{python}
# Load the pandas library
import pandas as pd
# Load numpy for array manipulation
import numpy as np
# Load seaborn plotting library
import seaborn as sns
import matplotlib.pyplot as plt

# Set font sizes in plots
sns.set(font_scale = 1.2)
# Display all columns
pd.set_option('display.max_columns', None)

Default = pd.read_csv("../data/Default.csv")
Default
```

```{python}
# Numerical summaries
Default.describe()
```
Graphical summary:
```{python}
#| eval: true
# Graphical summaries
plt.figure()
sns.pairplot(data = Default);
plt.show()
```



:::


## Initial split into test and non-test sets

It is a good idea to keep the proportions of `default = 'Yes'` to be roughly same between test and non-test data.

::: {.panel-tabset}

#### R

```{r}
##| eval: false
# For reproducibility
set.seed(425)
data_split <- initial_split(
  Wage, 
  # # stratify by percentiles
  # strata = "Salary", 
  prop = 0.75
  )

Wage_other <- training(data_split)
dim(Wage_other)
Wage_test <- testing(data_split)
dim(Wage_test)
```

#### Python

```{python}
from sklearn.model_selection import train_test_split

Default_other, Default_test = train_test_split(
  Default, 
  train_size = 0.75,
  random_state = 425, # seed
  stratify = Default.default
  )
Default_test.shape
Default_other.shape
```

Separate $X$ and $y$.
```{python}
# Non-test X and y
X_other = Default_other[['balance', 'income', 'student']]
y_other = Default_other.default
# Test X and y
X_test = Default_test[['balance', 'income', 'student']]
y_test = Default_test.default
```



:::

## Recipe (R) or Preprocessing (Python) 

::: {.panel-tabset}



#### R

```{r}
##| eval: false
norm_recipe <- 
  recipe(
    wage ~ year + age + education, 
    data = Wage_other
  ) %>%
  # create traditional dummy variables
  step_dummy(all_nominal()) %>%
  # zero-variance filter
  step_zv(all_predictors()) %>% 
  # B-splines of age
  step_bs(age, deg_free = 5) %>%
  # B-splines of year
  step_bs(year, deg_free = 4) %>%
  # center and scale numeric data
  step_normalize(all_predictors()) %>%
  # estimate the means and standard deviations
  prep(training = Wage_other, retain = TRUE)
norm_recipe
```

#### Python

Pre-processor for one-hot coding of categorical variables and then standardizing all numeric predictors.
```{python}
from sklearn.preprocessing import OneHotEncoder, StandardScaler, SplineTransformer
from sklearn.compose import ColumnTransformer
from sklearn.pipeline import Pipeline

# Transformer for numeric variables
numeric_tf = Pipeline(steps = [
  # ("scalar", StandardScaler()),
  ("bspline", SplineTransformer(degree = 3, extrapolation = 'linear'))
])
# Transformer for categorical variables
categorical_tf = Pipeline(steps = [
  ("encoder", OneHotEncoder(drop = 'first'))
])

# Column transformer
col_tf = ColumnTransformer(transformers = [
  ('num', numeric_tf, ['balance', 'income']),
  ('cat', categorical_tf, ['student'])
])
```

:::

## Model

Let's skip this step for now, because the model (or classifer) itself is being tuned by CV.

::: {.panel-tabset}


#### R

```{r}
##| eval: false
enet_mod <- 
  # mixture = 0 (ridge), mixture = 1 (lasso)
  linear_reg(penalty = tune(), mixture = tune()) %>% 
  set_engine("glmnet")
enet_mod
```

:::

## Workflow (R) or Pipeline (Python)

Here we bundle the preprocessing step (Python) or recipe (R) and model. Again remember KNN is a placeholder here. Later we will choose the better classifier according to cross validation.

::: {.panel-tabset}




#### R

```{r}
##| eval: false
lr_wf <- 
  workflow() %>%
  add_model(enet_mod) %>%
  add_recipe(norm_recipe)
lr_wf
```

#### Python

```{python}
from sklearn.neighbors import KNeighborsClassifier

pipe = Pipeline(steps = [
  ("col_tf", col_tf),
  ("model", KNeighborsClassifier())
  ])
pipe
```

:::

## Tuning grid

Set up the 2D grid for tuning.

::: {.panel-tabset}



#### R

```{r}
##| eval: false
param_grid <-grid_regular(
  penalty(range = c(-5, 0), trans = log10_trans()), 
  mixture(range = c(0, 1)),
  levels = c(penalty = 50, mixture = 6)
  )
param_grid
```

#### Python

```{python}
from sklearn.linear_model import LogisticRegression
from sklearn.discriminant_analysis import LinearDiscriminantAnalysis, QuadraticDiscriminantAnalysis
from sklearn.naive_bayes import GaussianNB

# Tune hyper-parameter(s)
nknots_grid = np.array(range(2, 6))
classifers = [
  LogisticRegression(),
  LinearDiscriminantAnalysis(),
  QuadraticDiscriminantAnalysis(),
  GaussianNB(),
  KNeighborsClassifier(n_neighbors = 5)
  ]
tuned_parameters = {
  "col_tf__num__bspline__n_knots": nknots_grid,
  "model": classifers
  }
tuned_parameters  
```

:::

## Cross-validation (CV)

::: {.panel-tabset}


#### R

Set cross-validation partitions.
```{r}
##| eval: false
set.seed(250)
folds <- vfold_cv(Wage_other, v = 10)
folds
```

Fit cross-validation.
```{r}
##| eval: false
enet_fit <- 
  lr_wf %>%
  tune_grid(
    resamples = folds,
    grid = param_grid,
    )
enet_fit
```

Visualize CV criterion.
```{r}
##| eval: false
enet_fit %>%
  collect_metrics() %>%
  print(width = Inf) %>%
  filter(.metric == "rmse") %>%
  ggplot(mapping = aes(x = penalty, y = mean)) + 
  geom_point() + 
  geom_line(aes(group = mixture)) + 
  labs(x = "Penalty", y = "CV RMSE") + 
  scale_x_log10(labels = scales::label_number())
```

Show the top 5 models ($\lambda$ values)
```{r}
##| eval: false
enet_fit %>%
  show_best("rmse")
```

Let's select the best model
```{r}
##| eval: false
best_enet <- enet_fit %>%
  select_best("rmse")
best_enet
```

#### Python

Set up CV partitions and CV criterion. Again it's a good idea to keep the case proportions roughly same between splits. According to the `GridSearchCV` [documentation](https://scikit-learn.org/stable/modules/generated/sklearn.model_selection.GridSearchCV.html), `StratifiedKFold` is used automatically.

```{python}
from sklearn.model_selection import GridSearchCV

# Set up CV
n_folds = 10
search = GridSearchCV(
  pipe,
  tuned_parameters,
  cv = n_folds, 
  scoring = "roc_auc",
  # Refit the best model on the whole data set
  refit = True
  )
```

Fit CV. This is typically the most time-consuming step.
```{python}
# Fit CV
search.fit(X_other, y_other)
```

Visualize CV results.
```{python}
#| eval: true
#| code-fold: true
cv_res = pd.DataFrame({
  "nknots": np.array(search.cv_results_["param_col_tf__num__bspline__n_knots"]),
  "classifier": np.array(search.cv_results_["param_model"]),
  "aucroc": search.cv_results_["mean_test_score"]
  })

plt.figure()
sns.relplot(
  kind = "line",
  data = cv_res,
  x = "nknots",
  y = "aucroc",
  hue = "classifier"
  ).set(
    xlabel = "Number of Knots",
    ylabel = "CV AUC"
);
plt.show()
```

Best CV AUC-ROC:
```{python}
search.best_score_
```


:::

## Finalize our model

Now we are done tuning. Finally, let’s fit this final model to the whole training data and use our test data to estimate the model performance we expect to see with new data.

::: {.panel-tabset}


#### R

```{r}
##| eval: false
# Final workflow
final_wf <- lr_wf %>%
  finalize_workflow(best_enet)
final_wf

# Fit the whole training set, then predict the test cases
final_fit <- 
  final_wf %>%
  last_fit(data_split)
final_fit

# Test metrics
final_fit %>% collect_metrics()
```

#### Python

Since we called `GridSearchCV` with `refit = True`, the best model fit on the whole non-test data is readily available.
```{python}
search.best_estimator_
```

The final AUC on the test set is
```{python}
from sklearn.metrics import roc_auc_score

roc_auc_score(
  y_test, 
  search.best_estimator_.predict_proba(X_test)[:, 1]
  )
```

:::